Are historical stage records useful to decrease the uncertainty of flood frequency analysis ? A 200-year long case study

### 2022-12-21 ArticleChProp/main.tex
retours Ĵérôme

\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage[pdftex]{graphicx}
\usepackage{tabularx}

%bib options
\usepackage[backend=bibtex,style=authoryear,natbib=true]{biblatex}
\addbibresource{Chap1.bib}

\usepackage{dsfont}
\usepackage{multirow}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{subcaption}
\usepackage{authblk}

%hyperlinks options
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    citecolor=cyan
    }

@@

    Rating curve uncertainty is also a major issue when dealing with streamflow series. Transforming stage into discharge requires calibration data (gaugings) to establish the stage/discharge relationship. Gaugings uncertainty depends on the measurement method (\citet{lecoz_quantification_2014}b and \citet{puechberty_charte_2017}). Moreover, the rating curve is also affected by uncertainties coming from the imperfection of the chosen model to represent the actual hydraulic configuration and from parameter estimation. Many methods have been proposed to quantify these uncertainties (\citet{petersen-overleir_bayesian_2009}; \citet{juston_rating_2014}; \citet{le_coz_combining_2014}; \citet{morlot_dynamic_2014}; \citet{coxon_novel_2015}; \citet{mcmillan_rating_2015}; \citet{mansanarez_rapid_2019}). A comparison of several of these methods has been recently proposed by \citet{kiang_comparison_2018}. Another important issue affecting streamflow data accuracy is rating changes. The stage/discharge relationship is frequently affected by changes coming from various origins that can be natural or anthropic, for instance: bed geometry evolution during floods or river works, aquatic vegetation growth and decay, ice cover... A regular monitoring through gaugings is essential to detect those changes \citep{ibbitt_gauging_1987}, that can be transient or sudden. Several methods have been proposed to deal with rating changes: estimating rating curves on moving temporal windows (\citet{westerberg_stage-discharge_2011} and \citet{guerrero_temporal_2012}), computing as many rating curves as there are gaugings \citep{morlot_dynamic_2014}, exploring changes in the annual minimum stages \citep{lapuszek_methods_2015}, selecting 0.5-years return period discharge as a threshold for rating changes \citep{mcmillan_impacts_2010}. More recently, \citet{darienzo_detection_2021} proposed a method based on a recursive segmentation procedure, accounting for both gaugings and rating curve uncertainties. This method has a particular interest when dealing with old and thus uncertain gaugings. Following the detection of rating shifts, rating curves should be estimated for each stability period. This task may not be straightforward, as the number of gaugings available within a stability period is not always sufficient to properly estimate the stage/discharge relationship for the whole discharge range. A common way to address this problem is to artificially repeat some gaugings from other stability periods (\citet{mcmillan_benchmarking_2012}; \citet{puechberty_charte_2017}). \citet{mansanarez_shift_2019} proposed an alternative approach to deal with this issue. They developed a stage-period-discharge (SPD) model where some parameters of the rating curve vary across periods, while the others are supposed constant. This method has the advantage of transferring information between periods to improve the rating curve estimation for all the stable stage/discharge periods, even when few gaugings are available.

@@

    \begin{figure}[h!]
    \centering
        \includegraphics[width = 8cm]{Figs/1-uTotSchema.png}
        \caption{Block diagram of the uncertainty propagation procedure. Grey blocks represents data, blue blocks stands for analysis methods/models.}
        \label{fig:ChProp}
    \end{figure}

@@

    \begin{figure}[h]
        \centering
        \includegraphics[width = 12cm]{Figs/5-Quant5perc_both.png}
        \caption{Time series of annual 5\% stage quantile of both Pont de Beaucaire (1816-1967) and Beaucaire Restitution (1970-2020) stations used to specify priors on rating shift parameters.}
        \label{fig:quantile5_both}
    \end{figure}

@@

    \begin{table}[h]
    \centering
    \resizebox{\columnwidth}{!}{%
    \begin{tabular}[c]{|c|c|c|c|c|cc|}
    \hline
    \multirow{2}{*}{\textbf{Date}} &
      \multirow{2}{*}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_1}$: gauge\\ reading\end{tabular}}} &
      \multirow{2}{*}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_2}$: instr.  \\ precision\end{tabular}}} &
      \multirow{2}{*}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_3}$: sensor\\ calibration\end{tabular}}} &
      \multirow{2}{*}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_4}$: datum\\ reference\end{tabular}}} &
      \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_5}$:\\ measurement frequency\end{tabular}}} \\ \cline{6-7} 
     &
       &
       &
       &
       &
      \multicolumn{1}{c|}{\textbf{Stage\textless5m}} &
      \textbf{Stage$\geq$5m} \\ \hline
    \textbf{Before 1840} &
       $\mathcal{N}(0,0.05)$ &
      - &
      - &
       $\mathcal{N}(0,0.06)$ &
      \multicolumn{2}{c|}{$\exp(2.18)$} \\ \hline
    \textbf{1840 - 1967} &
      $\mathcal{N}(0,0.05)$ &
      - &
      - &
       $\mathcal{N}(0,0.06)$ &
      \multicolumn{1}{c|}{\begin{tabular}[c]{@{}c@{}}$\exp(8.86)$\end{tabular}} &
      - \\ \hline
    \textbf{1970 - 2020} &
       - &
       $\mathcal{N}(0, 0.01/\sqrt{3})$ &
       $\mathcal{N}(0,0.06)$ &
      - &
      \multicolumn{2}{c|}{-} \\ \hline
    \end{tabular}%
    }
    \caption{Distributions used for the different sources of stage errors (in meters). $\mathcal{N}$ and $\exp$ respectively represent Gaussian and Exponential distributions.}
    \label{tab:StageErr}
    \end{table}

@@

         \begin{table}[ht]
            \centering
                \begin{tabular}{| l | l |} 
                        \hline
                        \textbf{Gauging method} & \textbf{Uncertainty standard deviation} \\
                        \hline
                        Current meter at 0.6 h and surface & 5\% \\
                        \hline
                        Current meter point by point & 3.5\% \\
                        \hline
                        Surface current meter & 7.5\% \\
                        \hline
                        Unknown type & 7.5\% \\
                        \hline
                        ADCP & 3.5\% \\
                        \hline
                        Floats before 1936 & 10\% \\
                        \hline
                        Hydrotachymeter before 1936 & 10\%\\
                        \hline
                \end{tabular}
            \caption{Gaugings uncertainty depending on the method used (hypotheses from \citet{bard_actualisation_2018}). Expressed as standard deviations of the measured discharge in \%.}
            \label{TabIcJau}
        \end{table}

@@

       The GEV distribution estimation procedure described in section \ref{sec:STOODS} is applied to the 205-year long AMAX discharges series at Beaucaire accounting for uncertainties. Vague priors are given to GEV parameters : flat priors for location and scale parameters, and Gaussian with zero mean and 0.2 standard deviation for the shape parameter.
       Flood quantiles results (figure \ref{subfig:GEV205y}) show that streamflow uncertainty dominates for the lowest return periods, but sampling uncertainty is taking over when the return period tends toward 1000 years (see figure \ref{subfig:UKplot4cases} bottom right  for a better understanding of the respective part of each source of uncertainty in this 205 years case). The AMAX observed discharges display a large variability of streamflow uncertainties. The three largest floods of this 205 years sample (1840, 1856 and 2003, by chronological order and from the most precise to the most uncertain) illustrate this point. Thus, not considering 1840 and 1856 years could have a large effect on the estimation of the maxpost quantiles value, as well as the uncertainty bound values. This is explored next by varying the sample size.

@@

        \caption{(a) Flood quantiles at Beaucaire. Error bars represent the AMAX observed discharges with 95\% streamflow uncertainty.
        (b) Quantiles estimation of three return periods for four sample sizes. 95\% total uncertainty envelop is represented by the yellow error bars. 
        (c) Contribution of both sources of uncertainty to the total uncertainty for four sample size. 
        (d) GEV shape parameter distribution considering streamflow or total uncertainty for four sample sizes. (e) 1000-year flood estimations for various sample sizes. Grey dotted lines are representing important changes in the sample data (the change from Pont de Beaucaire to Beaucaire Restitution gauge en 1967, the inclusion of 1856 flood and the inclusion of 1840 flood to the sample.}
        \label{fig:Quantiles}
        \end{figure}

@@

    \subsection{Limitations and potential improvements of the method}
    \paragraph{}
    The streamflow uncertainty analysis procedure proposed in this work is affected by several limitations. For simplicity purposes, gaugings stage uncertainty is assumed negligible, which is inconsistent with the consideration of stage measurement errors of section \ref{sec:StageErr}. \citet{horner_impact_2018} proposed a model that accounts for this source of uncertainty in a streamflow uncertainty propagation procedure. The gauging segmentation model proposed by \citet{darienzo_detection_2021} is an efficient tool to detect change points in the stage/discharge relationship, but is limited when dealing with old gauges for which gaugings are often scarce. Selecting a date within the posterior pdf is dubious when very few information are available about rating changes and channel morphology. Additional tests could be used to investigate about stage/discharge relationship, such as the analysis of flood recessions and morphogenic events occurrences proposed by \citet{darienzo_detection_2021-1}. Another limitation of the method comes from the elicitation of hydraulic priors in historical context, for which information about hydraulic configuration is scarce. This lack of knowledge is impacting when considering gauges where works inside the main channel were frequent. The Rhône River is part of those, even if it holds the interest of many historians who retrieved many pieces of hydraulic and morphological data. As described by \citet{petersen-overleir_accounting_2009}, this definitely has an impact on flood quantile estimates as the extrapolation of rating curves in this context is more uncertain. Another drawback comes from the fact that, for simplification purposes, we considered the days of annual maximum discharge as the days of annual maximum stage. For a given year, if $h_{j} > h_{k}$, it is not guaranteed that $Q_{j} > Q_{k}$ if there is an important rating change during this given year. The BaRatin \citep{lecoz_quantification_2014} based propagation model choice may also have an important impact on uncertainty estimation, as the seven rating curve uncertainty estimation methods compared by \citet{kiang_comparison_2018} concluded in a very large differences in uncertainty intervals.

@@

\newpage
\printbibliography
\section{Annexes}

>>>>

% \usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage[pdftex]{graphicx}
\usepackage{tabularx}

%bib options
\usepackage[backend=biber,style=authoryear,bibstyle=authoryear,natbib=true,
giveninits=true,
uniquename=false,
uniquelist=false,
% firstinits=false,
maxcitenames=2,
date=year, maxbibnames=99,url=false]{biblatex}
% \usepackage[authoryear]{natbib}
\addbibresource{Chap1.bib}
\usepackage{dsfont}
\usepackage{multirow}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{subcaption}
\usepackage{authblk}
%hyperlinks options
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    citecolor=cyan
    }

@@

    Rating curve uncertainty is also a major issue when dealing with streamflow series. Transforming stage into discharge requires calibration data (gaugings) to establish the stage/discharge relationship. Gaugings uncertainty depends on the measurement method (\citet{lecoz_quantification_2014} and \citet{puechberty_charte_2017}). Moreover, the rating curve is also affected by uncertainties coming from the imperfection of the chosen model to represent the actual hydraulic configuration and from parameter estimation. Many methods have been proposed to quantify these uncertainties (\citet{petersen-overleir_bayesian_2009}; \citet{juston_rating_2014}; \citet{le_coz_combining_2014}; \citet{morlot_dynamic_2014}; \citet{coxon_novel_2015}; \citet{mcmillan_rating_2015}; \citet{mansanarez_rapid_2019}). A comparison of several of these methods has been recently proposed by \citet{kiang_comparison_2018}. Another important issue affecting streamflow data accuracy is rating changes. The stage/discharge relationship is frequently affected by changes coming from various origins that can be natural or anthropic, for instance: bed geometry evolution during floods or river works, aquatic vegetation growth and decay, ice cover... A regular monitoring through gaugings is essential to detect those changes \citep{ibbitt_gauging_1987}, that can be transient or sudden. Several methods have been proposed to deal with rating changes: estimating rating curves on moving temporal windows (\citet{westerberg_stage-discharge_2011} and \citet{guerrero_temporal_2012}), computing as many rating curves as there are gaugings \citep{morlot_dynamic_2014}, exploring changes in the annual minimum stages \citep{lapuszek_methods_2015}, selecting 0.5-years return period discharge as a threshold for rating changes \citep{mcmillan_impacts_2010}. More recently, \citet{darienzo_detection_2021} proposed a method based on a recursive segmentation procedure, accounting for both gaugings and rating curve uncertainties. This method has a particular interest when dealing with old and thus uncertain gaugings. Following the detection of rating shifts, rating curves should be estimated for each stability period. This task may not be straightforward, as the number of gaugings available within a stability period is not always sufficient to properly estimate the stage/discharge relationship for the whole discharge range. A common way to address this problem is to artificially repeat some gaugings from other stability periods (\citet{mcmillan_benchmarking_2012}; \citet{puechberty_charte_2017}). \citet{mansanarez_shift_2019} proposed an alternative approach to deal with this issue. They developed a stage-period-discharge (SPD) model where some parameters of the rating curve vary across periods, while the others are supposed constant. This method has the advantage of transferring information between periods to improve the rating curve estimation for all the stable stage/discharge periods, even when few gaugings are available.

@@

    \begin{figure}[h!]
    \centering
        \includegraphics[width = 8cm]{Figs/1-uTotSchema.png}
        \caption{Block diagram of the uncertainty propagation procedure. Grey blocks represent data, blue blocks stand for analysis methods/models that correspond to the sub-sections of this article. $h$ is the water stage, $uh$ the stage uncertainty, $Q$ the discharge, $uQ$ the discharge uncertainty, $t$ is the time and $RP$ the return period of flood quantiles.}
        \label{fig:ChProp}
    \end{figure}

@@

    \begin{figure}[h]
        \centering
        \includegraphics[width = 15cm]{Figs/5-Quant5perc_both.png}
        \caption{Time series of annual 5\% stage quantile of both Pont de Beaucaire (1816-1967) and Beaucaire Restitution (1970-2020) stations used to specify priors on rating shift parameters.}
        \label{fig:quantile5_both}
    \end{figure}

@@

    \begin{table}[h]
    \centering
    \resizebox{\columnwidth}{!}{%
    \begin{tabular}[c]{|c|c|c|c|c|cc|}
    \hline
    \multirow{2}{*}{\textbf{Date}} &
      \multirow{2}{*}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_1}$: gauge\\ reading\end{tabular}}} &
      \multirow{2}{*}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_2}$: instr.  \\ precision\end{tabular}}} &
      \multirow{2}{*}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_3}$: sensor\\ calibration\end{tabular}}} &
      \multirow{2}{*}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_4}$: datum\\ reference\end{tabular}}} &
      \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}$\boldsymbol{\delta_5}$:\\ measurement frequency\end{tabular}}} \\ \cline{6-7} 
     &
       &
       &
       &
       &
      \multicolumn{1}{c|}{\textbf{Stage\textless5m}} &
      \textbf{Stage$\geq$5m} \\ \hline
    \textbf{Before 1840} &
       $\mathcal{N}(0,0.05)$ &
      - &
      - &
       $\mathcal{N}(0,0.06)$ &
      \multicolumn{2}{c|}{$\exp(2.18)$} \\ \hline
    \textbf{1840 - 1967} &
      $\mathcal{N}(0,0.05)$ &
      - &
      - &
       $\mathcal{N}(0,0.06)$ &
      \multicolumn{1}{c|}{\begin{tabular}[c]{@{}c@{}}$\exp(8.86)$\end{tabular}} &
      - \\ \hline
    \textbf{1970 - 2020} &
       - &
       $\mathcal{N}(0, 0.01/\sqrt{3})$ &
       $\mathcal{N}(0,0.06)$ &
      - &
      \multicolumn{2}{c|}{-} \\ \hline
    \end{tabular}%
    }
    \caption{Distributions used for the different sources of stage errors (in meters). $\mathcal{N}$ and $\exp$ represent Gaussian and Exponential distributions respectively.}
    \label{tab:StageErr}
    \end{table}

@@

         \begin{table}[ht]
            \centering
                \begin{tabular}{| l | l |} 
                        \hline
                        \textbf{Gauging method} & \textbf{Standard uncertainty} \\
                        \hline
                        Current meter at 0.6 h and surface & 5\% \\
                        \hline
                        Current meter point by point & 3.5\% \\
                        \hline
                        Surface current meter & 7.5\% \\
                        \hline
                        Unknown type & 7.5\% \\
                        \hline
                        ADCP & 3.5\% \\
                        \hline
                        Floats before 1936 & 10\% \\
                        \hline
                        Hydrotachymeter before 1936 & 10\%\\
                        \hline
                \end{tabular}
            \caption{Gaugings uncertainty depending on the method used (hypotheses from \citet{bard_actualisation_2018}). Expressed as standard deviations of the measured discharge in \%.}
            \label{TabIcJau}
        \end{table}

@@

       The GEV distribution estimation procedure described in section \ref{sec:STOODS} is applied to the 205-year long AMAX discharges series at Beaucaire accounting for uncertainties. Vague priors are given to GEV parameters : flat priors for location and scale parameters, and Gaussian with zero mean and 0.2 standard deviation for the shape parameter. This shape parameter prior is coherent with \citet{martins_generalized_2000} suggestions.
       Flood quantiles results (figure \ref{subfig:GEV205y}) show that streamflow uncertainty dominates for the lowest return periods, but sampling uncertainty is taking over when the return period tends toward 1000 years (see figure \ref{subfig:UKplot4cases} bottom right  for a better understanding of the respective part of each source of uncertainty in this 205 years case). The AMAX observed discharges display a large variability of streamflow uncertainties. The three largest floods of this 205 years sample (1840, 1856 and 2003, by chronological order and from the most precise to the most uncertain) illustrate this point. Thus, not considering 1840 and 1856 years could have a large effect on the estimation of the maxpost quantiles value, as well as the uncertainty bound values. This is explored next by varying the sample size.

@@

        \caption{FFA results of the Rhône River at Beaucaire. (a) Fitted GEV distribution on the full sample (1816-2020). Error bars represent the AMAX observed discharges with 95\% streamflow uncertainty.
        (b) Quantiles estimation of three return periods for four sample sizes. 95\% total uncertainty envelop is represented by the yellow error bars. 
        (c) Contribution of streamflow and sampling uncertainties to the total uncertainty for four sample sizes. 
        (d) GEV shape parameter distributions considering streamflow or total uncertainty for four sample sizes. (e) 1000-year flood estimations for various sample sizes. Grey dotted lines are representing important changes in the sample data (the change from Pont de Beaucaire to Beaucaire Restitution gauge en 1967, the inclusion of 1856 flood and the inclusion of 1840 flood to the sample.}
        \label{fig:Quantiles}
        \end{figure}

@@

    \subsection{Limitations and potential improvements of the method}
    \paragraph{}
    The streamflow uncertainty analysis procedure proposed in this work is affected by several limitations. For simplicity purposes, gaugings stage uncertainty is assumed negligible, which is inconsistent with the consideration of stage measurement errors of section \ref{sec:StageErr}. \citet{horner_impact_2018} proposed a model that accounts for this source of uncertainty in a streamflow uncertainty propagation procedure. The gauging segmentation model proposed by \citet{darienzo_detection_2021} is an efficient tool to detect change points in the stage/discharge relationship, but is limited when dealing with old gauges for which gaugings are often scarce. Selecting a date within the posterior pdf is dubious when very few information are available about rating changes and channel morphology. Additional tests could be used to investigate about stage/discharge relationship, such as the analysis of flood recessions and morphogenic events occurrences proposed by \citet{darienzo_detection_2021-1}. Another limitation of the method comes from the elicitation of hydraulic priors in historical context, for which information about hydraulic configuration is scarce. This lack of knowledge is impacting when considering gauges where works inside the main channel were frequent. The Rhône River is part of those, even if it holds the interest of many historians who retrieved many pieces of hydraulic and morphological data. As described by \citet{petersen-overleir_accounting_2009}, this definitely has an impact on flood quantile estimates as the extrapolation of rating curves in this context is more uncertain. Another drawback comes from the fact that, for simplification purposes, we considered the days of annual maximum discharge as the days of annual maximum stage. For a given year, if $h_{j} > h_{k}$, it is not guaranteed that $Q_{j} > Q_{k}$ if there is an important rating change during this given year. The BaRatin \citep{le_coz_combining_2014} based propagation model choice may also have an important impact on uncertainty estimation, as the seven rating curve uncertainty estimation methods compared by \citet{kiang_comparison_2018} concluded in a very large differences in uncertainty intervals.

@@

\newpage
\printbibliography
\newpage

\section{Annexes}
